Composition-climate paper

### 2016-08-24 Composition_Ecology.Rmd
Cleaned up the references.  Finished substantial text edits.

These compounding and confounding interactions between historic climate change and land use, and the consequent shifts in realized climate space have important implications for the management and genetics of forest communities.  On the one hand, the observation that historical realized niches are broader than modern niches suggests that some tree genera may have more resilience to climate change.  Critically, the combination of regional land-use change and climate change based on contemporary data alone leads to the under-prediction of species resilience to climate change regionally. Regionally, annual temperature is expected to increase in Wisconsin by 3.6^o^C in 2055 from a 1980s baseline [@wicci2011climate]. Every tree genera examined in this study had historical $T_{max}$ values between 0.4 and 1.3^o^C higher than modern estimates, and with $P_{ann}$ between 42 and 86mm lower than modern estimates. Projected changes in $P_{ann}$ range from -17 - 110mm from a 1980 baseline to 2055 based on fourteen global climate models compiled as part of the Climate Model Intercomparison Project [@wicci2011climate]. Higher temperatures and lower precipitation across taxa distributions during the pre-settlement period indicate the potential for greater resilience to climate change than might be assessed using modern data only.  This effect may be critical for species that reach their northern or western range limits in the upper Midwestern United States (e.g., *Tsuga*, *Fagus*), because a narrowing of the apparent environmental envelope, due to the twin action of land-use change and changing climate, may result in underprediction of species resilience to climate change.

However, projected climate change far exceeds the shifts due to historical land use and climate change.  In addition, coupled migration and adaptation in response to future climate change is likely to be impeded by the extirpation of communities due to land use change and loss of genetic diversity.  Species migration and genetic adaptation are coupled across time-scales [@davis2001range].  The loss of forest communities along the forest-prairie boundary in the upper Midwest is particularly important given that the direction of climate change regionally is towards warmer conditions, conditions that exists at the forest-prairie boundary.  The loss of genotypes adapted to warmer and drier conditions (the lagging edge of distributions for many taxa in the Upper Midwest) may mean more rapid losses at the lagging edges of the species range.  Long distance gene flow of adaptive traits to central and leading populations from the lagging edge [@hu2006seed] is lost when land use conversion to agricultural production results in the extirpation of locally adapted populations.  Thus net gene flow in populations along the remaining "lagging" edge will be from the central and leading populations, individuals maladapted to warmer conditions. This maladaptive gene flow could potentially amplify the effects of climate change on populations along the forest-prairie boundary leading to increasing risk of local extinction among natural populations [@kremer2012long].

Species distribution models fall along a continuum from more correlative to more process-based [@dormann2012correlation], and correlational species-distribution models are likely to be the most affected by historical land-use change and the sampling design of modern survey data. However, even process based models are partially parameterized from contemporary distributions, particularly with respect to the environmental tolerances of plant functional types [*e.g.*, @wullschleger2014plant].  Spatially biased land-use conversion could result in the extirpation of well-adapted genotypes within a population. In this case, both correlative and process based models might be affected by spatially biased land-use conversion.

In summary, historical datasets and multitemporal studies such as this oneadd to our ability to understand species realized climate niches and their adaptive potential in the face of climate change. They also serve to underscore the important interactions between climate change and land use change at a population level.  The strong changes in climate and land use over the past two centuries, combined with the likelihood that tree distributions are lagging recent climate changes [@svenning2013disequilibrium], create the strong likelihood that tree-climate distributions have shifted and that current tree-climate distributions are not representative of pre-settlement conditions.  Given the importance of modern distributional datasets for predicting species range shifts, compounding may result in the impression of narrower climatic tolerances than actually exist, leading to underestimates of species resilience.  At the same time, confounding interactions between historic climate change and land-use effects may increase predictive uncertainty.

>>>>

These compounding and confounding interactions between historic climate change and land use, and the consequent shifts in realized climate space have important implications for the management and genetics of forest communities.  On the one hand, the observation that historical realized niches are broader than modern niches suggests that some tree genera may have more resilience to climate change than expected based on contemporary observations alone. Regionally, annual temperature is expected to increase in Wisconsin by 3.6^o^C in 2055 from a 1980s baseline [@wicci2011climate] and $P_{ann}$ projections range from -17 to 110mm by 2055 relative to 1980s baselines, based on fourteen global climate models from the Climate Model Intercomparison Project [@wicci2011climate]. Every tree genera examined in this study had historical $T_{max}$ values from 0.4 to 1.3^o^C higher than modern estimates, and $P_{ann}$ from 42 to 86mm lower than modern estimates. These higher temperatures and lower precipitation across taxa distributions during the pre-settlement period indicate the potential for greater resilience to climate change than might be assessed using modern data only.  This evidence for historical resilience to warmer and dryer climates may be particularly critical for species that reach their northern or western range limits in the upper Midwest (e.g., *Tsuga*, *Fagus*).

However, projected temperature rises by 2100 far exceed the shifts due to historical land use and climate change.  In addition, coupled migration and adaptation in response to future climate change is likely to be impeded by the extirpation of communities due to land use change and loss of genetic diversity. Local adaptation is an important component of tree responses to climate change [@aitken2008adaptation]. Hence, the loss of genotypes adapted to warmer and drier conditions (the trailing edge of distributions for many taxa in the Upper Midwest) caused by land use may mean more rapid losses at the trailing edges of the species range.  Long distance gene flow of adaptive traits to central and leading populations from the trailing edge [@hu2006seed] is lost when land use conversion to agricultural production results in the extirpation of locally adapted populations.  Thus net gene flow in populations along the remaining trailing edge will be from the central and leading populations, individuals maladapted to warmer conditions. This maladaptive gene flow could potentially amplify the effects of climate change on populations along the forest-prairie boundary leading to increasing risk of local extinction among natural populations [@kremer2012long].

Species distribution models fall along a continuum from more correlative to more process-based [@dormann2012correlation], and correlational species-distribution models are likely to be the most affected by the use of contemporary observational datasets that are biased by historical climate and land-use change and climatic disequilibrium. However, even process based models are partially parameterized from contemporary distributions, particularly with respect to the environmental tolerances of plant functional types [*e.g.*, @wullschleger2014plant]. In this case, both correlative and process based models might be affected by the shifting climatic niches demonstrated here, and the underlying processes of spatially biased land-use conversion, changing climates, and climatic disequilibrium.

In summary, historical datasets and multitemporal studies such as this oneadd to our ability to understand the stability and dynamic of species-climate distributions and their adaptive potential in the face of climate change.  The strong changes in climate and land use over the past two centuries, combined with the likelihood that tree distributions are lagging recent climate changes [@svenning2013disequilibrium], create the strong likelihood that tree-climate distributions have shifted and that current tree-climate distributions are not representative of pre-settlement conditions.  Given the importance of modern distributional datasets for predicting species range shifts, compounding interactions between historical climate and land use change may result in the impression of narrower climatic tolerances than actually exist, leading to underestimates of species resilience. Both empirical and process-based vegetation models need to combine, whenever possible, both contemporary and historical information on tree distributions, tree-climate relationships, and their respective shifts through time.

Acknowledgements
---------------------
The authors would like to thanks the large number of individuals who have worked to first, undertake the Public Lands System Survey, to bring the original survey data together, to digitize and standardize much of the survey results, and finally, to assist in interpreting and compiling the data in its present form. This material is based upon work carried out by the PalEON Project (http://paleonproject.org), under NSF Award Numbers 1241868 and 1065656.  Maps were made using data from Natural Earth (http://naturalearth.com), which provides raster & vector map data in the public domain.
